---
title: "Demonstration of R Markdown"
author: "Wenhao Jiang"
date: "2024-09-06"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

You will need to create R Markdown files to submit assignments. 

* You can combine both **text** and **codes** in a single Markdown file.
* Lab materials can be found from [my github page](https://github.com/wenhaojiangsoc/statslab).

You can change line by adding `two spaces`, or typing `\`, or adding `<br>`, at the end of the first line.

### Typing Equations

You can type equations in Markdown using a LaTeX-style syntax.<br>  
For example, the sample mean (statistic) is given by $\hat{\mu} = \frac{\sum_{i = 1}^{n}y_i}{n}$\
You can also type the equation in a separate line by having two `$$` at the beginning and the end of the equation. This makes equations more clear.
$$
\hat{\mu} = \frac{\sum_{i = 1}^{n}y_i}{n}
$$
Notations and syntax can be strange at the beginning. You will get used to it gradually. You can refer to this [page](https://rpruim.github.io/s341/S19/from-class/MathinRmd.html) for more information.

### Integrating Codes

```{r}
## your codes

```

```{r install packages, message=FALSE, eval=FALSE}
## install package individually
install.packages("tidyverse")

## or install multiple packages together
install.packages(c("tidyverse", "gridExtra", "kableExtra"))
```

```{r load packages, message=FALSE}
## load packages
library(tidyverse)
library(gridExtra)
library(kableExtra)
```

```{r data types, message=FALSE}

```

```{r clean and summarize data, message=FALSE}

```

```{r misc, message=FALSE}

```


